---
title: 'CMTH 642 Data Analytics: Advanced Methods'
output:
  word_document: default
  pdf_document: default
  html_document: default
---
<center> <h1> Assignment 1 (10%) </h1> </center>
<center>  <h3> Ana Abesamis </h2> </center>
<center> <h3> CMTH642 DHB ID# 501148675 </h2> </center>
---

#### 1. Read the csv files in the folder. (4 points)
```{r}
macronutrients_df <- read.csv(file="USDA_Macronutrients.csv",header=TRUE, sep=",")
micronutrients_df <- read.csv(file="USDA_Micronutrients.csv",header=TRUE, sep=",")
head(macronutrients_df)
head(micronutrients_df)
```

#### 2. Merge the data frames using the variable "ID". Name the Merged Data Frame "USDA". (4 points)
```{r}
USDA <- merge(macronutrients_df,micronutrients_df, by="ID")
head(USDA)
```

#### 3. Check the datatypes of the attributes. Delete the commas in the Sodium and Potasium records. Assign Sodium and Potasium as numeric data types. (4 points)
```{r}
str(USDA)
USDA$Sodium <- gsub("\\,","",USDA$Sodium)
USDA$Sodium <- as.numeric(USDA$Sodium)
USDA$Potassium <- gsub("\\,","",USDA$Potassium)
USDA$Potassium <- as.numeric(USDA$Potassium)

str(USDA)
```

#### 4. Remove records (rows) with missing values in more than 6 attributes (columns). How many records remain in the data frame? (4 points)
```{r}
clean_USDA <- USDA[rowSums(is.na(USDA))<7,]
str(clean_USDA)

# Without records with >6 missing values, we are left with 6965 observations of 15 variables.
```

#### 5. For records with missing values for Sugar, Vitamin E and Vitamin D, replace missing values with mean value for the respective variable. (4 points)
```{r}
mean_sugar <- mean(clean_USDA$Sugar,na.rm=TRUE)
mean_vite <- mean(clean_USDA$VitaminE,na.rm=TRUE)
mean_vitd <- mean(clean_USDA$VitaminD,na.rm=TRUE)
clean_USDA$Sugar[is.na(clean_USDA$Sugar)]<-mean_sugar
clean_USDA$VitaminE[is.na(clean_USDA$VitaminE)]<-mean_vite
clean_USDA$VitaminD[is.na(clean_USDA$VitaminD)]<-mean_vitd
```

#### 6. With a single line of code, remove all remaining records with missing values. Name the new Data Frame "USDAclean". How many records remain in the data frame? (4 points)
```{r}
USDAclean <- clean_USDA[complete.cases(clean_USDA),]
str(USDAclean)

# Without records with any missing values, we are left with 6310 observations of 15 variables.
```

#### 7. Which food has the highest sodium level? (4 points)
```{r}
USDAclean$Description[which.max(USDAclean$Sodium)]
```

#### 8. Create a histogram of Vitamin C distribution in foods. (4 points)
```{r}
hist(USDAclean$VitaminC, breaks=200, freq=TRUE,
     main="Histogram of Vitamin C Distribution in Foods",
     col="orange",xlim=c(0,200), xlab="mg")
```

#### 9. Create one boxplot to illustrate the distribution of values for TotalFat, Protein and Carbohydrate. (4 points)
```{r}
par(mfrow=c(1,3))
boxplot(USDAclean$TotalFat)
boxplot(USDAclean$Protein)
boxplot(USDAclean$Carbohydrate)
```

#### 10. Create a scatterplot to illustrate the relationship between a food's TotalFat content and its Calorie content. (4 points)
```{r}
plot(USDAclean$TotalFat,USDAclean$Calories,
     main="Relationship between Total Fat and Calorie Content",
     xlab="Total Fat",ylab="Calories", col="lightgreen")
```

#### 11. Add a variable to the data frame that takes value 1 if the food has higher sodium than average, 0 otherwise. Call this variable HighSodium. Do the same for High Calories, High Protein, High Sugar, and High Fat. How many foods have both high sodium and high fat? (4 points)
```{r}
USDAclean$HighSodium <- ifelse(USDAclean$Sodium>mean(USDAclean$Sodium,na.rm=TRUE),1,0)
USDAclean$HighCalories <- ifelse(USDAclean$Calories>mean(USDAclean$Calories,na.rm=TRUE),1,0)
USDAclean$HighProtein <- ifelse(USDAclean$Protein>mean(USDAclean$Protein,na.rm=TRUE),1,0)
USDAclean$HighSugar <- ifelse(USDAclean$Sugar>mean(USDAclean$Sugar,na.rm=TRUE),1,0)
USDAclean$HighFat <- ifelse(USDAclean$TotalFat>mean(USDAclean$TotalFat,na.rm=TRUE),1,0)
head(USDAclean,10)

sum(USDAclean$HighSodium==1 & USDAclean$HighFat==1)
```

#### 12. Calculate the average amount of iron, for high and low protein foods. (4 points)
```{r}
hp_iron <- mean(USDAclean$Iron[USDAclean$HighProtein==1],na.rm=TRUE)
lp_iron <- mean(USDAclean$Iron[USDAclean$HighProtein==0],na.rm=TRUE)
```

#### 13. Create a function for a "HealthCheck" program to detect unhealthy foods. Use the algorithm flowchart below as a basis. (4 points)
```{r}
require(jpeg)
img<-readJPEG("HealthCheck.jpg")
plot(1:4, ty = 'n', ann = F, xaxt = 'n', yaxt = 'n')
rasterImage(img,1,1,4,4)
```
```{r}
HealthCheck_fun <- function(Sodium,Sugar,Fat){
  ifelse(Sodium==1,
         ifelse(Sugar==1,
                ifelse(Fat==1,"Fail","Pass"),
            "Pass"),
         "Pass")
}
```

#### 14. Add a new variable called HealthCheck to the data frame using the output of the function. (4 points)
```{r}
USDAclean$HealthCheck <- HealthCheck_fun(USDAclean$HighSodium,USDAclean$HighSugar,USDAclean$HighFat)
head(USDAclean$HealthCheck,10)
```

#### 15. How many foods in the USDAclean data frame fail the HealthCheck? (4 points)
```{r}
sum(USDAclean$HealthCheck=="Fail")
```

#### 16. Visualize the correlation among Calories, Protein, Total Fat, Carbohydrate, Sodium and Cholesterol. (4 points)
```{r}
# Created a subset of the attributes to correlate from the clean dataframe
sub <- c("Calories","Protein","TotalFat","Carbohydrate","Sodium","Cholesterol")
subUSDA <- USDAclean[,sub]
head(subUSDA)

plot(subUSDA)

# The scatterplot diagram gives a simple yet comprehensive visual of relationships
# among multiple attributes. In this case, we get a view of the correlations
# between each of the 6 given attributes. We can easily see positive correlation
# (shown by the upward trend of Calories v TotalFat) as well as negative
# correlations (shown by the downward trends of Protein v TotalFat, TotalFat v
# Carbohydrates, etc)

```

#### 17. Is the correlation between Calories and Total Fat statistically significant? Why? (4 points)
```{r}
cor.test(USDAclean$Calories,USDAclean$TotalFat)
# Yes, the correlation between Calories and Total Fat is statistically significant. According
# to the Pearson correlation test, they have a correlation of 0.804495 with a p-value of 2.2e-16.
# Because the p-value is very small and is less than 0.05, we can say this correlation is
# significant.
```

#### 18. Create a Linear Regression Model, using Calories as the dependent variable Protein, Total Fat, Carbohydrate, Sodium and Cholesterol as the independent variables. (4 points)
```{r}
model <- lm(Calories~Protein+TotalFat+Carbohydrate+Sodium+Cholesterol,data=USDAclean)
```

#### 19. Which independent variable is the least significant? Why? (4 points)
```{r}
summary(model)

# The least significant attribute would be Sodium, based on the p-value of the linear regression
# model. If we want a 95% confidence interval, then alpha must not be greater than 0.05, which
# is the case for all coefficients but Sodium.
```

#### 20. Create a new model by using only the significant independent variables. (4 points)
```{r}
new_model <- lm(Calories~Protein+TotalFat+Carbohydrate+Cholesterol,data=USDAclean)
summary(new_model)
```

#### 21. A new product is just produced with the following data: Protein=0.1, TotalFat=37, Carbohydrate=400, Cholesterol=75, Sugar=NA, Calcium=35, Iron=NA, Potassium=35, VitaminC=10,	VitaminE=NA, VitaminD=NA. Based on the new model you created, what is the predicted value for Calories? (4 points)
```{r}
# Using the model cal = 3.989679*Protein + 8.771904*TotalFat + 3.743859*Carbohydrate
#                       + 0.010980*Cholesterol + 4.077907
predicted_cal <- 3.989679*0.1 + 8.771904*37 + 3.743859*400 + 0.010980*75 + 4.077907
predicted_cal
```

#### 22. If the Carbohydrate amount increases from 400 to 40000 (10000% increase), how much change will occur on Calories in percent? Explain why? (4 points)
```{r}
predicted_cal2 <- 3.989679*0.1 + 8.771904*37 + 3.743859*40000 + 0.010980*75 + 4.077907
predicted_cal2
change <- (predicted_cal2-predicted_cal)/predicted_cal
change

# A 10000% increase in Carbohydrates would result to a 8112.972% increase in Calories. This is
# because the Carbohydrate coefficient is a positive number. With each increase in Carbohydrates,
# the value of Calories also increase.
```

#### 23. Prepare an exploratory data analysis question (without using predictions) about the dataset. Write a code to answer your question. Visualise your answer. (Your question should be related to at least three attributes) (12 points)
```{r}

# EDA Question: Of all the given macronutrients and micronutrients, which pairs
# have the strongest correlation?

# Why this question: This is an extension to question# 16 which I feel was not
# done fully. I think it's important to get the correlations among ALL
# attributes. It was one of the early EDA questions we covered in our CIND119
# final project. But a scatterplot will not be effective when we're dealing with
# 13 variables, so maybe a correlation matrix will help. This was taught by my
# CIND123 TA and at the time I thought it was the coolest visualization she 
# showed us :)

# Prepared the subset of the dataframe and checked the values
eda <- c("Calories","Protein","TotalFat","Carbohydrate","Sodium","Cholesterol",
         "Sugar","Calcium","Iron","Potassium","VitaminC","VitaminE","VitaminD")
edaUSDA <- USDAclean[,eda]
head(edaUSDA)

# Created a table of values for correlating all the attributes
cortableUSDA <- cor(edaUSDA)
cortableUSDA

# Used corrplot to get a fun visualization of correlations
# install.packages("corrplot")
library(corrplot)
corrplot.mixed(cortableUSDA, upper="ellipse", lower="number")

# According to the correlation matrix plot below, total fat and calories have
# the strongest positive correlation, while carbohydrate and sugar come as a
# close second. Without looking at the numbers of the correlation test, you
# can visualize that going towards a darker shade of blue indicates a positive
# correlation, while going the opposite way indicates a negative correlation.

# And with every correlation test, there is the famous last words: "correlation
# does not mean causation." This means that while Total Fat has the strongest
# correlation with Calories, we do not have enough evidence to say that high
# Total Fat leads to more calories consumed.

```

This is the end of Assignment 1

Ceni Babaoglu, PhD
